Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-02-27 paper.tex
update

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.85]{beramono}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor,colortbl}
\usepackage{url}
\usepackage{listings}
\usepackage{paralist}
\usepackage{wrapfig}
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{flushend}

@@

\subsection{Motivation}

@@

@@

We use Scala language to demonstrate the idea and each step of transformations.
There are two main reasons to do so:
1) The code does not diminish the accuracy of the material
than formal and mathematical notations, 
which are heavily used in other papers in static analysis or semantics.
The Scala code in this paper can be easily back-translated into formal notations.
2) The code is also executable with only few changes, which make it particularly fit for 
presenting topic: syntactical transformations of programs.

\section{Background}

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine (AAM) is a systematic methodology that derives sound 
abstract interpreters for higher-order functional languages from concrete 
abstract machines \cite{van2012systematic, van2010abstracting}. 
An abstracting abstract machine implements computable abstract semantics that 
approximates the runtime behaviors of programs.
Since the state space of concrete execution is possibly infinite,
so to analyze programs, the key insight of AAM approach is to allocate both bindings
and continuations on the store, and then bound the addresses space to be finite.
Thus the approximated machine-state space is finite, and computable.

In this section, we derive the abstracting abstract machine from concrete
CESK machines, and also show that how to instantiate
useful $k$-call-sensitive analysis.

@@

Similar with CESK machines, the machine state of the AAM has a control expression,
an environment, a store and continuation, and in addition with a time-stamp.
But there are several notable differences of store between AAMs and concrete CESK machines.
In AAM, the store maps addresses to sets of values, so it stores all possible values
for a particular address. Thus the dereference an address becomes non-deterministic.
Also, the store performs a join rather than overwriting when updating elements of the store.
Third, the continuations are likewise allocated on store instead of formed a linked list,
and the continuation in a state then turns into a continuation address.

@@

The structure of function @step@ is similar with concrete CESK machines, 
except non-determinism which makes @step@ returns a list of reachable 
successor states.
In the beginning of @step@, we call function @tick@ to update 
the time.
Then we have two cases to consider: 1) Current control expression is @Let@,
and @App(f, ae)@ is being bind to @x@.
Fristly we retrieve the set of closures that @f@ may represents.
For each closure in the set, we do almost the same as what we did in concrete 
CESK machines. But note that since the continuation is allocated on store (@kstore@),
a new continuation address @new_kaddr@ is constructed, and a new frame
@Frame(x, e, env)@ paired with the current continuation address @kaddr@
is joined into this address. Finally, a list of states is generated.

@@

To run (analyze) a program, as CESK machine, we use @inject@ to construct
the initial state given a program. 
Note that the initial continuation store has a special mapping that maps continuation address
@Halt@ to empty set of continuations.
We also provide an empty program contour as initial time.
\begin{lstlisting}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{lstlisting}

@@

Pushdown control-flow is a property in analysis that precisely models
the runtime call-stack of the analyzed program. 
A pushdown control-flow analysis provides as exact as runtime return-flow 
when analyzing the program, but traditional control-flow analysis collapse 
the state space using finite space abstraction thus cause imprecise stack modeling.

To see the problem of spurious return-flows, we can consider the following example:

@@

In traditional $k$-CFA algorithm and the AAM we showed in section \ref{aam}, 
the return-flow of invocation @(id 2)@
is going to both call-sites @(id 1)@ and @(id 2)@. Therefore, the
returned value @2@ for variable @y@ is also propagated to 
variable @x@ and then imprecise analysis result is arose.
This return-flow merging is inevitable even we increase the context-sensitivity.
If we use the monovariant analysis (i.e. 0CFA), the analyzing result would be @x@ and 
@y@ point to set @{1, 2}@, because the algorithm not distinguishes
@z@ comes from different call-sites.
Under $1$-CFA, the algorithm is able to distinguish that variable @z@ of function
@id@ has two different values at two call-sites, so variable @y@ 
would not be polluted by @1@, but two call-sites still share the same continuation and
it is still helplessness for distinguish return-flows.
TODO: do need more details to explain this example? (as in P4F paper)

@@

A naive solution to is to use precise call stack modeling as we did in concrete
CESK machine. Instead of allocating continuations on store and recording an 
address of continuations, we leave the call stack unabstracted.

In the definition of state @State@, continuation store disappears, and
@konts@ becomes a list of frames. Other components keep the same.

@@

The state transition function @step@ is still non-determinism, but
the only non-determinism happens when dereferencing the callee @f@ 
from the store at call-site @App(f, ae)@. 
For each closure of @f@ in the set, we do exactly
the same as what we did in concrete CESK machines: a new frame is constructed
and pushed onto the stack.
In the second case, since the continuation @konts@ is a list of 
frames, we can simply decompose it to differentiate an empty stack
or a frame exists on the top of stack. 
If the stack is empty, then empty successors is returned, because the computation
is finished.
Otherwise, we matched a frame @Frame(x, e, env_f)@. 
This transition is deterministic, because we only have one frame, this frame
is then popped and the new state uses the rest of stack.

@@

In recent years, there are great efforts \cite{vardoulakis2010cfa2, earl2012introspective, 
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return 
match based on small-step AAM which is what we described in section \ref{aam}.

@@

\textbf{Pushdown for Free.}
\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machine that is both easy to implement and
computational inexpensive.
The only change in code is when allocating a continuation address, not only
keep track of the entry-point expression of callee, but also hold the target environment.
The target environment is the environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the rest things remain the same.

@@

TODO: explain the reason why it works.
Notably, this change only causes constant-factor increase in time complexity of analysis.

@@

On the other side, with the insight from Reynold's seminal paper 
\emph{Definitional interpreters for higher-order programming languages} \cite{reynolds1972definitional}
that the defined language will inherit properties
from the defining language, \citeauthor{darais2017abstracting} construct abstracting 
definitional interpreters (ADI) that automatically inherit the pushdown control-flow 
property from its defining language.

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[scaled=0.85]{beramono}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor,colortbl}
\usepackage{url}
\usepackage{listings}
\usepackage{paralist}
\usepackage{wrapfig}
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{flushend}
\usepackage{tikz}

\usetikzlibrary{arrows,chains,matrix,positioning,scopes}
\makeatletter
\tikzset{join/.code=\tikzset{after node path={%
\ifx\tikzchainprevious\pgfutil@empty\else(\tikzchainprevious)%
edge[every join]#1(\tikzchaincurrent)\fi}}}
\makeatother

\tikzset{>=stealth',every on chain/.append style={join},
         every join/.style={->}}
\tikzstyle{labeled}=[execute at begin node=$\scriptstyle,
   execute at end node=$]

@@

@@

The paper is organized as follows. 
TODO: add a diagraph on transformations.
Section \ref{background} reviews the background.

\begin{tikzpicture}
  \matrix (m) [matrix of math nodes, row sep=3em, column sep=3em]
    { abstracting abstract machines & A  & B  & C  & 0 \\
      abstract machines & A' & B' & C' & 0 \\ };
  { [start chain] \chainin (m-1-1);
    \chainin (m-1-2);
    { [start branch=A] \chainin (m-2-2)
        [join={node[right,labeled] {\eta_1}}];}
    \chainin (m-1-3) [join={node[above,labeled] {\varphi}}];
    { [start branch=B] \chainin (m-2-3)
        [join={node[right,labeled] {\eta_2}}];}
    \chainin (m-1-4) [join={node[above,labeled] {\psi}}];
    { [start branch=C] \chainin (m-2-4)
        [join={node[right,labeled] {\eta_3}}];}
    \chainin (m-1-5); }
  { [start chain] \chainin (m-2-1);
    \chainin (m-2-2);
    \chainin (m-2-3) [join={node[above,labeled] {\varphi'}}];
    \chainin (m-2-4) [join={node[above,labeled] {\psi'}}];
    \chainin (m-2-5); }
\end{tikzpicture}

@@

We use Scala language to demonstrate the idea and each step of transformations.
The reader should have moderate familiarity to Scala's syntax, such as
case classes, pattern matching and for comprehension.

There are two main reasons to use a real-world language:
1) The code does not diminish the accuracy of the material
than formal and mathematical notations, 
which are heavily used in other papers in static analysis or semantics.
The Scala code in this paper can be easily back-translated into formal notations.
2) As a functional pearl, the code in this paper is executable with only few changes, 
which make it particularly fit for presenting syntactical transformations on abstract
interpreters.

\section{Background} \label{background}

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine (AAM) is a systematic methodology that derives sound 
abstract interpreters for higher-order functional languages from concrete 
abstract machines \cite{van2012systematic, van2010abstracting}. 
An abstracting abstract machine implements computable abstract semantics that 
approximates the runtime behaviors of programs.
Since the state space of concrete execution is possibly infinite,
so to analyze programs, the key insight of AAM approach is to allocate both bindings
and continuations on the store, and then bound the addresses space to be finite.
Since each component of state is finite, 
thus the abstracted machine-state space is finite, and computable.

In this section, we derive the abstracting abstract machine from concrete
CESK machines, and also show that how to instantiate
useful $k$-call-sensitive control-flow analysis.

@@

Similar with CESK machines, the machine state of AAM has a control expression,
an environment, a store and continuation, and in addition with a time-stamp.
But there are several notable differences of store between AAMs and concrete CESK machines.
In AAM, the store maps addresses to sets of values, so it stores all possible values
for a particular address. Thus the dereference an address becomes non-deterministic.
Also, the store performs a join rather than overwriting when updating elements of the store.
Third, the continuations are likewise allocated on store instead of formed a linked list,
and the continuation in a state then turns into a continuation address.

@@

The structure of function @step@ is similar with concrete CESK machines, 
except non-determinism which makes @step@ returns a list of reachable 
successor states.
In the beginning of @step@, we call function @tick@ to update 
the time.
Then we have two cases to consider: 1) Current control expression is @Let@,
and @App(f, ae)@ is being bind to @x@.
Fristly we retrieve the set of closures that @f@ may represents.
For each closure in the set, we do almost the same as what we did in concrete 
CESK machines. But note that since the continuation is allocated on store (@kstore@),
a new continuation address @new_kaddr@ is constructed, and a new frame
@Frame(x, e, env)@ paired with the current continuation address @kaddr@
is joined into @new_kaddr@. Finally, a list of successors states is generated.

@@

To run (analyze) a program, as CESK machine, we use function @inject@ to construct
the initial state given a program. 
Note that the initial continuation store has a built-in mapping that maps continuation address
@Halt@ to empty set of continuations.
We also provide an empty program contour as initial time.
\begin{lstlisting}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{lstlisting}

@@

Pushdown control-flow is a property in analysis that precisely models
the runtime call-stack of the analyzed program. 
A pushdown control-flow analysis provides as exact as runtime return-flow 
when analyzing the program, but traditional control-flow analysis collapses 
the state space into a finite space thus causes imprecise stack modeling.

To see how traditional CFA suffers from spurious return-flows, 
we can consider the following example:

@@

In traditional $k$-CFA algorithm and the AAM we showed in section \ref{aam}, 
the return-flow of invocation @(id 2)@
is going to both call-sites @(id 1)@ and @(id 2)@. Therefore, the
returned value @2@ for variable @y@ is also propagated to 
variable @x@ and then imprecise analysis result is arose.
This return-flow merging is inevitable even we increase the context-sensitivity.
If we use the monovariant analysis (i.e. 0CFA), the analyzing result would be @x@ and 
@y@ point to set @{1, 2}@, because the algorithm not distinguishes
@z@ comes from different call-sites.
Under $1$-CFA, the algorithm is able to distinguish that variable @z@ of function
@id@ has two different values at two call-sites, so variable @y@ 
would not be polluted by @1@. 
But two call-sites still share the same continuation and
it is still helplessness for separating the return-flows.
TODO: do need more details to explain this example? (as in P4F paper)

@@

A naive solution to is to use precise call stack as we did in concrete
CESK machine. Instead of allocating continuations on store and recording an 
address of continuations in state, we leave the call stack unabstracted.

In the definition of state @State@, the continuation store disappears, and
component @konts@ becomes a list of frames. Other components keep the same.

@@

The state transition function @step@ is still non-deterministic, but
the only non-determinism happens when dereferencing the callee @f@ 
from the store at call-site @App(f, ae)@. 
For each closure of @f@ in the set, we do exactly
the same as what we did in concrete CESK machines: a new frame is constructed
and pushed onto the stack.
In the second case, since the continuation @konts@ is a list of 
frames, we can simply decompose it to differentiate an empty stack case
or a frame exists on the top of stack. 
If the stack is empty, then empty list of successors is returned, 
because the computation is finished.
Otherwise, we matched a frame @Frame(x, e, env_f)@ on the top of stack.
The frame is then poped, and the new state uses the rest of stack.
Since we can only access one frame at a time, this transition is deterministic.

@@

In recent years, there are significant efforts \cite{vardoulakis2010cfa2, earl2012introspective, 
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return 
match based on small-step AAM which is what we described in section \ref{aam}.

@@

\textbf{Pushdown for Free.}
\citeauthor{gilray2016pushdown} proposed a continuation address allocator for 
small-step abstracting abstract machines that is both simple to implement and
computational inexpensive.
The only change in code is when allocating a continuation address, not only
keep track of the entry-point expression of callee, but also hold the target environment.
The target environment is the binding environment extended from closure's environment
(namely @val new_env = env_c + (v -> baddr)@). All the rest things remain the same.

@@

TODO: explain the reason why it works.

Notably, this change only causes constant-factor increase in time complexity of analysis
if the store is widened.

@@

On the other side, with the insight from Reynold's seminal paper 
\emph{Definitional interpreters for higher-order programming languages} \cite{reynolds1972definitional}
that the defined language will inherit properties
from the defining language, \citeauthor{darais2017abstracting} construct abstracting 
definitional interpreters that automatically inherit the pushdown control-flow 
property from its defining language.
